---
title: "Crab Sattelites"
author: "Sean Steele"
date: "12/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lmtest)
library(margins)
library(sandwich)
library(mfx)
```

#Read data and adjust varaibles 
```{r}
crabs <- read.table("http://www.stat.ufl.edu/~aa/cat/data/Crabs.dat", header=TRUE)
#Create clearer color variables
crabs$light<-as.factor(crabs$color==1)
crabs$medium<-as.factor(crabs$color==2)
crabs$mDark<-as.factor(crabs$color==3)
crabs$dark<-as.factor(crabs$color==4)
```

#Explore data with plots
```{r}
#Color and width seem to show a possible trend in number of mate satellites heavy concentration of zero satellites hints that a binary response framework (no sat/ vs sat ~ width and/or weight) as best indicator for a general trend
plot(x= crabs$width, y = crabs$sat) 
plot(x= crabs$weight, y = crabs$sat)
plot(x= crabs$color, y = crabs$sat)

#explore data as binary response plots
plot(x= crabs$width, y= crabs$y)
plot(x= crabs$weight, y= crabs$y)

#Explore histogram of number of satellites 
hist(crabs$sat, breaks= 30) #histogram hints that satellites could be count variable with a Poission distribution -- explore after logit model
```

#explore basic linear probability model
```{r}
summary(crabs)
#colorless LPM model
LPM1 <- lm(y~ width + weight, data = crabs)
summary(LPM1)
coeftest(LPM1, type = "HC1") #Hetroskedasticity robust standard errors required
#Weight variable is insignificant at even the 10% alpha level, but the model F statistic indicates the overall model is signficant

#Drop Weight from LPM model
LPM2 <- lm(y~ width, data = crabs)
summary(LPM2)
coeftest(LPM2, type = "HC1")
#Width now highly significant with a one centimeter increase in width corresponding to a ~9% increase in having a satellite

#Add nominal scale color variables with dark color as the base group
LPM3 <- lm(y~width + light + medium + mDark, data = crabs)
summary(LPM3)
coeftest(LPM3, type = "HC1")
#linear probability model indicates that medium and medium dark colors are significantly (at 5% alpha level) more likely to have a satellite than dark colored crabs by 29.3% and 23.5% respectively. 

```

#Fit the logit and probit model for a more appropriate binary response model
```{r}
#estimate logit with color included
logit1 <- glm(y~width + light + medium + mDark, family = binomial(link = "logit"), data = crabs)
summary(logit1) #AIC = 197.46
#obtain APEs
summary(margins(logit1))
#each one centimeter increase in width corresponds to an average marginal effect of ~8% increase in having satellites while light, medium, and dark colored crabs have a 20.5%, 18%, and 26% increase in probability of having satellites compared to dark crabs 

#see if probit model has better fit to data
probit1 <- glm(y~width + light + medium + mDark, family = binomial(link = "probit"), data = crabs)
summary(probit1) #210.38 -- logit model better fit to data

#plot logit model
logit_plot <- ggplot(crabs, aes(x=width, y=y)) +
  geom_point(alpha = 0.3) +
  stat_smooth(method = "glm", se= TRUE, method.args = list(family=binomial))
logit_plot
```

#Determine affect of colors on satellite presence by likihood ratio test
```{r}
#unrestricted model is logit1: Pr[y=1| width, light,medium,medium dark]
#restricted model: Pr[y=1|width]

#unrestricted residual deviance
URdeviance<-logit1$deviance

#restricted model
logit2<-glm(y~width, family = binomial(link = "logit"), data=crabs)
#restricted residual deviance
Rdeviance<-logit2$deviance

#LR test stat
LR<-(Rdeviance-URdeviance)

#test with chi square under 3 degrees of freedom
pchisq(LR, df=3, lower.tail=FALSE)
#At 5% significance level must reject that color as a factor variable is significant

```

#Does treating color as a numeric ordinal variable fit the data better and retain significance?
```{r}
logit3 <- glm(y ~ width + color, family = binomial(link = "logit"), data = crabs)
summary(margins(logit3))
#The average marginal effect of width remains nearly the same at about ~8% increase of satellite per one centimeter increase in width, and a one unit increase in darkness (ie: light to medium light or medium to dark) results in a ~9% decrease in probability of having satellites and is significant at the 5% level

#check significance 
#unrestricted model is logit1: Pr[y=1| width, color]
#restricted model: Pr[y=1|width]

#unrestricted residual deviance
URdeviance2<-logit3$deviance

#restricted still logit 2 from above
#LR test stat
LR2<-(Rdeviance-URdeviance2)

#test with chi square under 1 degree of freedom
pchisq(LR2, df=1, lower.tail=FALSE)
#model is significant at 5% level (p= 0.021)

#check model fitness between ordinal and factor variable
BIC(logit1) #213.2
BIC(logit3) #204.6

#Ordinal variable with number of regressor penalty (BIC) shows better fit to data in addition to higher significance 
```

#Confirm weight is still insignificant with color as an ordinal variable
```{r}
#unrestricted model is logit1: Pr[y=1| width, color, weight]
#restricted model: Pr[y=1|width,color] (logit3)

logit4 <- glm(y ~ width + color + weight, family = binomial(link = "logit"), data = crabs)
URdeviance3 <- logit4$deviance
Rdeviance2 <- logit3$deviance
LR3 <- (Rdeviance2 - URdeviance3)
#test with chi square under 1 degree of freedom
pchisq(LR3, df=1, lower.tail=FALSE)
#Weight still insignificant - certainly discard consideration at this point
```

#Explore poisson model for satelitte prediction
```{r}
pois_fit1 <- glm(sat ~ width + color, data = crabs, family ="poisson")
summary(pois_fit1)
#poisson model shows significance for both variables established in logit model with same direction

#determine size of the average marginal effect of width and color
poissonmfx(formula = sat ~ width + color, data = crabs, atmean = FALSE)
#AME for every 1 cm increase in width results in an average of 0.433 satellites, and every unit increase of darkness results in an average decrease of number satellites by 0.49

#Percentage change for satellite number when color may be more interpretable 
coef_pois1 <- coef(pois_fit1)
(exp(coef_pois1[3])-1)*100
# ~15.6% decrease in number of satellites as the crab color darkens one step
```

#Poisson model corrections - quasipoisson and negative binomial
```{r}
#Estimate dispersion parameter
pois1_resid <- crabs$sat - pois_fit1$fitted.values
sigma <- sqrt(sum(pois1_resid^2/pois_fit1$fitted.values)/(nrow(crabs)-3))
dispersion <- sigma^2
dispersion #dispersion is 3.22 indicating overdispersion

#correct standard errors for over dispersion with quasipossion model
quasi_pois1 <- glm(sat ~ width + color, data = crabs, family = "quasipoisson")
summary(quasi_pois1)
#width remains significant while color no longer does

#quasi poisson APE
mean(crabs$sat)*coef(quasi_pois1)[2]
#APE remains nearly the same under quasipossion model with a 1cm increase corresponding to 0.437 more  satellites on average

#Estimate by negative binomial and pick best fit model
negbin1 <- glm.nb(sat ~ width + color, data = crabs)
summary(negbin1)

#Find AME
negbinmfx(formula = negbin1, data=crabs, atmean = FALSE, robust = TRUE)

#One cm increase corresponds to an average of 0.53 more satellites 

BIC(pois_fit1)
BIC(negbin1) #negative binomial has better fit to data


```






